\documentclass[11pt, a4paper]{article}

% --- PREAMBLE ---
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{geometry}
\usepackage{listings} % For code blocks
\usepackage{color}
\usepackage{hyperref}
\usepackage{enumitem}
\usepackage{amsmath} % Added for math
\usepackage{amssymb} % Added for math

% Geometry
\geometry{top=2.5cm, bottom=2.5cm, left=2.5cm, right=2.5cm}

% Code styling
\definecolor{dkgreen}{rgb}{0,0.6,0}
\definecolor{gray}{rgb}{0.5,0.5,0.5}
\definecolor{mauve}{rgb}{0.58,0,0.82}

\lstset{frame=tb,
  language=R,
  aboveskip=3mm,
  belowskip=3mm,
  showstringspaces=false,
  columns=flexible,
  basicstyle={\small\ttfamily},
  numbers=none,
  numberstyle=\tiny\color{gray},
  keywordstyle=\color{blue},
  commentstyle=\color{dkgreen},
  stringstyle=\color{mauve},
  breaklines=true,
  breakatwhitespace=true,
  tabsize=3
}

% Title
\title{\textbf{Technical Specification: Algorithmic Implementation and Software Architecture of DyProp}}
\author{DyProp Development Team}
\date{\today}

\begin{document}

\maketitle

\section{Overview}
The \texttt{dyprop} package is an R/Bioconductor extension of the \texttt{propr} ecosystem. It implements the mathematical framework of Dynamic Proportionality using a high-performance C++ backend for vectorized template matching and an R frontend for S4 class management, event classification, and hierarchical validation.

\section{Data Transformation Pipeline}

\subsection{Zero Handling Protocol}
[cite_start]To enable log-ratio analysis on sparse genomic data without masking biological signals (e.g., gene silencing), we implement a rigorous two-step imputation strategy[cite: 140, 141].
\begin{enumerate}
    \item \textbf{Step A: Technical Correction (kNN Smoothing)}
    \begin{itemize}
        \item \textbf{Objective:} To correct technical dropouts (sporadic zeros) based on manifold proximity.
        [cite_start]\item \textbf{Implementation:} We apply k-Nearest Neighbor (kNN) pooling ($k=5$) in PCA space[cite: 142].
        \item \textbf{Logic:} If a zero is sporadic (neighbors have counts), it is smoothed. [cite_start]If a zero is consistent, it is preserved as a Biological Zero[cite: 143, 144].
    \end{itemize}
    
    \item \textbf{Step B: Biological Imputation (Bayesian)}
    \begin{itemize}
        [cite_start]\item \textbf{Objective:} To allow log-ratios involving Biological Zeros (e.g., silenced genes) to be computed[cite: 145].
        [cite_start]\item \textbf{Implementation:} We utilize the \texttt{zCompositions::cmultRepl} function[cite: 146].
        [cite_start]\item \textbf{Logic:} Zeros are replaced by a probabilistic estimate $\delta$ derived from the limit of detection, ensuring that an ``On/Off'' switch is detected as a large, quantifiable change in the log-ratio[cite: 147, 148].
    \end{itemize}
\end{enumerate}

\subsection{Coordinate Projection}
Following imputation, the data matrix $\mathbf{X}$ is projected into Euclidean space using the Centered Log-Ratio (CLR) transformation. To ensure robustness against outliers (e.g., PCR artifacts), we apply Winsorization at the ratio level:

\begin{lstlisting}
# R Pseudo-code
# 1. Full CLR Transform
clr_matrix <- propr::propr(imputed_counts)@logratio

# 2. Robust Ratio Calculation (e.g., Gene A vs Gene B)
ratio_vector <- clr_matrix[, A] - clr_matrix[, B]
clean_vector <- winsorize(ratio_vector, probs = c(0.01, 0.99))
\end{lstlisting}

\section{Algorithmic Engine: The Vectorized Scanner}

\subsection{Phase 1: Just-In-Time Dictionary Generation}
To bypass iterative regression, we pre-compute a dictionary of archetypal Boundary Functions. Upon initialization, the C++ backend generates a template matrix $\mathcal{M}$ ($K_{archetypes} \times T$) dynamically to match the user's pseudotime resolution. The grid resolution is coupled to the detection limit via the Nyquist criterion ($\Delta\tau \le \epsilon_{min}$).

\subsection{Phase 2: Vectorized Template Matching (C++)}
We perform a global search for topological tipping points using dense linear algebra.

\begin{itemize}
    \item \textbf{Input:} Ratio Matrix $\mathbf{Y}$ ($n \times p$), Template Matrix $\mathcal{M}$.
    \item \textbf{Operation:} We compute the convolution via matrix multiplication using \texttt{RcppArmadillo}.
\end{itemize}

\begin{lstlisting}[language=C++]
// src/dyprop.cpp (Snippet)
arma::mat scan_templates(arma::mat Y, arma::mat M) {
    // Y: [Pairs x Time], M: [Archetypes x Time]
    // Result: [Pairs x Archetypes] correlation matrix
    arma::mat scores = Y * M.t(); 
    return scores;
}
\end{lstlisting}

\subsection{Phase 3: Quadratic Interpolation}
To recover continuous parameter estimates from the discrete grid, we apply Parabolic Interpolation.
\begin{itemize}
    \item \textbf{Logic:} Given the best matching template index $i$, we fit a local parabola to scores $(R_{i-1}, R_i, R_{i+1})$.
    \item \textbf{Output:} The vertex of the parabola yields the precise continuous Tipping Point $\hat{\tau}$ and Sharpness $\hat{\epsilon}$.
\end{itemize}

\section{Software Architecture}

\subsection{Class Structure}
[cite_start]We define the \texttt{dyprop} S4 class to manage the complex outputs of dynamic analysis[cite: 159].

\begin{lstlisting}
setClass("dyprop",
    contains = "propr",       # Inherits basic CoDa slots from parent package
    slots = c(
        pseudotime = "numeric", # The ordered coordinate
        design = "matrix",      # Covariates matrix 
        
        # The "Movie" Data (3D Arrays)
        Phi_Array = "array",    # Dynamic Instability
        Rho_Array = "array",    # Dynamic Coupling
        
        # The "Catalog"
        events = "data.frame",  # Columns: GeneA, GeneB, Event_Class, Tau, Epsilon, FDR
        
        # Validation
        glmm_fits = "list",     # Storage for Phase 3 GLMM objects
        dictionary_meta = "list"
    )
)
\end{lstlisting}

\subsection{Core Methods}
\begin{description}
    \item[\texttt{prepareComposition(counts)}] Wrapper for the kNN + Imputation workflow. [cite_start]Returns clean CLR matrix[cite: 161].
    \item[\texttt{scanDynamics(object)}] Calls the C++ backend. [cite_start]Populates \texttt{Phi\_Array}, \texttt{Rho\_Array}, and performs the Template Match[cite: 162].
    \item[\texttt{classifyEvents(object)}] Applies the Multi-Evidence Decision Tree to label events as ``Switch'' or ``Decoupling''.
    \item[\texttt{validateGLMM(object)}] Runs the hierarchical validation module on significant candidates.
\end{description}

\section{Phase 4: Network Reconstruction & Differential Topology}

To translate statistical events into biological mechanisms, we implement a dedicated module for \textbf{Differential Topology Inference}. This allows users to map the specific regulatory edges that are rewired during the phase transition.

\subsection{Regime Definition (Temporal Slicing)}
For gene pairs identified as a \textbf{Stoichiometric Switch} (Type 1), we utilize the estimated Tipping Point $\hat{\tau}$ and Sharpness $\hat{\epsilon}$ to rigorously partition the trajectory into stable regimes. We explicitly exclude the ``Boundary Layer'' (the transition region) to prevent transition-state noise from contaminating the static estimates.

\begin{lstlisting}[language=R]
# R Logic: Define Stable Regimes
get_regimes <- function(pseudotime, tau, epsilon) {
    # Define the Boundary Layer width (e.g., 2*epsilon covers ~90% of transition)
    boundary_width <- 2 * epsilon
    
    # Partition cells
    idx_pre  <- which(pseudotime < (tau - boundary_width))  # Stable State A
    idx_post <- which(pseudotime > (tau + boundary_width))  # Stable State B
    
    return(list(pre = idx_pre, post = idx_post))
}
\end{lstlisting}

\subsection{Differential Topology ($\Delta \mathbf{P}$)}
We compute the static proportionality matrix $\mathbf{P}$ (the standard Rho metric from \texttt{propr}) separately for the Pre- and Post-transition cell subsets. We define the \textbf{Rewiring Matrix} $\mathbf{\Delta P}$ as:

$$ \mathbf{\Delta P} = \mathbf{P}_{\text{post}} - \mathbf{P}_{\text{pre}} $$

\begin{itemize}
    \item \textbf{Decoupling ($\Delta P_{ij} \ll 0$):} The regulatory constraint is relaxed. The gene pair moves from a coupled state to an independent state.
    \item \textbf{Coupling ($\Delta P_{ij} \gg 0$):} A new regulatory constraint is established. The gene pair locks into a fixed stoichiometric ratio.
\end{itemize}

\subsection{Haywire Hub Analysis}
We identify driver regulators by calculating the \textbf{Differential Degree Centrality} ($\Delta k$) for each gene. [cite_start]A high $\Delta k$ indicates a gene that has simultaneously rewired its relationship with many targets---a signature of a master regulator or a ``Haywire Hub''[cite: 32].

$$ \Delta k_i = \sum_{j \neq i} |\Delta P_{ij}| $$

\noindent\textbf{Downstream Integration:} The package exports the list of top $\Delta k$ genes to facilitate motif enrichment analysis (e.g., via \texttt{RCisTarget}), linking the detected stoichiometric switch to specific Transcription Factors (e.g., \textit{MYC}, \textit{TP53}).

\section{Phase 5: Hierarchical Validation Module}

To support clinical experimental designs, DyProp includes a confirmatory GLMM module utilizing \texttt{glmmTMB}. This framework is chosen over standard \texttt{lme4} to leverage the Student's t-distribution for outlier robustness and to explicitly model heteroscedasticity in pseudotime.

\begin{itemize}
    \item \textbf{Objective:} To separate biological trajectory dynamics from random batch/patient effects while controlling for heavy-tailed residual noise.
    \item \textbf{Input:} The subset of significant gene pairs identified in Phase 2.
    \item \textbf{Model:} Robust Generalized Linear Mixed Model.
    $$ y_{ijk} \sim \mathcal{T}(\mu_{ijk}, \sigma_{ij}, \nu) $$
    $$ \mu_{ijk} = \beta_0 + f(\text{Pseudotime}_{ij}) + \text{Condition}_k + (1 | \text{Patient}_k) $$
    $$ \log(\sigma_{ij}) = \gamma_0 + \gamma_1 \text{Pseudotime}_{ij} \quad \text{(Dispersion Model)} $$
\end{itemize}

\begin{lstlisting}[language=R]
# R Implementation Strategy
validate_pair <- function(data, pair_id) {
  # formula: Ratio ~ Spline(Time) + Random Intercept
  fit <- glmmTMB::glmmTMB(
    formula = log_ratio ~ splines::ns(pseudotime, df=3) + (1 | patient_id),
    dispformula = ~ pseudotime, # Models changing variance
    family = glmmTMB::t_family(), # Robust to outliers
    data = data
  )
  return(fit)
}
\end{lstlisting}



\end{document}