---
title: "Introduction to Dynamic Proportionality Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to Dynamic Proportionality Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

## Introduction

DyProp (Dynamic Proportionality) is a statistical engine designed to identify mechanisms of network rewiring in longitudinal genomic data (scRNA-seq, scDNA-seq, Multi-omics). It integrates **Compositional Data Analysis (CoDa)** with **Singular Perturbation Theory** to detect stoichiometric switches and network decoupling.

This vignette demonstrates the core workflow using a synthetic dataset.

## Setup

```{r setup}
library(dyprop)
library(glmmTMB) # For validation
```

## 1. Simulate Data

We generate a synthetic scRNA-seq count matrix with 100 cells and 20 genes along a pseudotime trajectory [0, 1].
We implant a **stoichiometric switch** between `Gene1` and `Gene2` at pseudotime $\tau = 0.5$.

```{r sim_data}
set.seed(42)
n_cells <- 100
n_genes <- 20
pseudotime <- sort(runif(n_cells, 0, 1))

# Initialize counts
counts <- matrix(rnbinom(n_cells * n_genes, mu = 10, size = 1), nrow = n_cells, ncol = n_genes)
colnames(counts) <- paste0("Gene", 1:n_genes)

# Implant Switch: Gene1 and Gene2 are proportional before t=0.5, then decoupled
tau <- 0.5
idx_pre <- which(pseudotime < tau)
idx_post <- which(pseudotime >= tau)

# Pre: High Correlation
counts[idx_pre, 1] <- rnbinom(length(idx_pre), mu = 50, size = 10)
counts[idx_pre, 2] <- counts[idx_pre, 1] + rnbinom(length(idx_pre), mu = 5, size = 10) # Correlated

# Post: Decoupled
counts[idx_post, 1] <- rnbinom(length(idx_post), mu = 50, size = 10)
counts[idx_post, 2] <- rnbinom(length(idx_post), mu = 10, size = 10) # Independent noise
```

## 2. Pre-processing

We use `prepareComposition` to perform:
1. **kNN Pooling** (correct technical dropouts, using GLM-PCA if available).
2. **Imputation** (handle biological zeros).
3. **CLR Transformation** (Center Log-Ratio).

```{r preprocess}
# Create dyprop object
# Input: Raw counts and pseudotime
dobj <- prepareComposition(counts, pseudotime, k = 5)

print(dobj)
```

## 3. Scan Dynamics

We run `scanDynamics` to identify candidate transition events. This uses a high-performance C++ engine to perform vectorized template matching (Sigmoid/Double-Sigmoid) against the log-ratio trajectories of all gene pairs.

```{r scan}
# Scan for candidates
# (Returns all pairs sorted by score; filtering happens downstream or manually)
dobj <- scanDynamics(dobj)

# Check detected events (Top 6)
head(dobj@events)
```

### 3.5 FDR Estimation (New)

We estimate the statistical significance of these events using an empirical Null Model (permuting pseudotime).

```{r fdr}
# Estimate FDR (using small permutations for vignette speed)
dobj <- estimateFDR(dobj, n_permutations = 1000)

# Filter for significant events (FDR < 0.1)
sig_events <- dobj@events[dobj@events$FDR < 0.1, ]
print(paste("Significant Events:", nrow(sig_events)))
```

## 4. Classify Events

The scanner identifies *potential* events. We now classify them into **Stoichiometric Switches** (Type 1) or **Network Decoupling** (Type 2) using mechanistic metrics:
*   $\Phi$ (Dynamic Instability)
*   $\rho$ (Dynamic Coupling)

```{r classify}
dobj <- classifyEvents(dobj)

# View classifications
table(dobj@events$Event_Class)
```

## 5. Network Reconstruction

For confirmed transition events, we analyze the topological rewiring.

### Differential Proportionality
We calculate the change in the proportionality matrix $\Delta \mathbf{P}$.

```{r network}
# Analyze the primary switch pair (Gene1 vs Gene2)
# Ideally found in dobj@events, but for demo we pick explicity if not found (due to random noise)
if (nrow(dobj@events) > 0) {
    geneA <- dobj@events$GeneA[1]
    geneB <- dobj@events$GeneB[1]
    tau_est <- dobj@events$Tau[1]
    eps_est <- dobj@events$Epsilon[1]

    rewiring <- calcRewiring(dobj, geneA, geneB, tau = tau_est, epsilon = eps_est)

    # Check Delta P for the partner
    # Index of partner in the vector
    idx_B_cols <- which(colnames(dobj@logratio) == geneB)
    print(paste("Delta P (Gene A vs B):", rewiring$Delta_P[idx_B_cols]))
}
```

## 6. Validation (GLMM)

Finally, we validate significant events using a Generalized Linear Mixed Model (GLMM) to distinguish biological signals from random effects (if patient metadata were available).

```{r validate}
# In this synthetic example, we don't have patient metadata, so it uses a simple LM fallback.
dobj <- validateGLMM(dobj)

# Inspect model fits
if (length(dobj@glmm_fits) > 0) {
    # Fits are stored as a list
    # names(dobj@glmm_fits)
    # print(dobj@glmm_fits[[1]])
}
```
